---
title: "Keystone exercise: Estuary Fish Survey (Data rescue)"
author: "Kendra Sturdee"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
 html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 6
    fig_width: 10
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
# Set options, housekeeping
knitr::opts_chunk$set(
	echo = TRUE,
	include = TRUE,
	message = FALSE,
	warning = FALSE)
```

# Introduction

The goal of this exercise is to wrangle extremely messy data, and produce a final clean dataset and figure on the Ross River estuary fish surveys. The data are from a previous PhD student who conducted surveys over 30 days in May 2026 at four different sites along the Ross River estuary, Townsville, Queensland, Australia. The data include catch counts for four different fish species, as well as environmental data (temperature, salinity, and turbidity) collected at each site using sonde sensors.

# Housekeeping

```{r housekeeping}
#| message: true
#| warning: false
rm(list = ls())
```

# Packages

```{r packages}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(lubridate)
library(readxl)
```

# Load files

# Main dataset

The catch data is split across 4 tabs in one excel file Load. We must load them indivudally.

```{r load-catch-1}
catch1 <- read_excel(here::here("data/workshop2/estuary_catch_log.xlsx"), sheet = "Lower_ross")
catch2 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 2)
catch3 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 3)
catch4 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 4)
```

# Inspect individual tabs
Must inspect the data for missing values, discreptancies, and errors before we proceed. 
```{r inspect-catch-tabs}
catch1
catch2
catch3
catch4
```
The species names are messy but they are consistent across the four tabs. The site names are also messy and inconsistent across the four tabs.
# Bind rows together
```{r bind-catch}
catch_data <- bind_rows(catch1, catch2, catch3, catch4)
# Clean up the environment
rm(catch1, catch2, catch3, catch4) # Clean up the environment
```

# Preliminary checks
```{r inspect-catch}
catch_data
head(catch_data)
glimpse(catch_data)
str(catch_data)
```

From this inspection we cannot see how many sites and species we have in the dataset. We can check this by using distinct() or by converting to factors

```{r distinct-catch}
catch_data |> 
  distinct(site)

catch_data |> 
  distinct(species)
```
Now I know I have four sites, and four species. By naming convention, I can infer that the sites are located along the Ross River estuary, but I need to check this against the metadata once I load it. The species names are also messy and will need to be cleaned before I can join with the taxonomic dictionary to get scientific names.

# Load metadata
```{r load-metadata}
metadata <- read_csv("../data/workshop2/estuary_metadata.csv")
```

# Checks
```{r inspect-metadata}
head(metadata)
glimpse(metadata)
```
Sites names differ between the catch log and the metadata. The catch log has site names in the format "lower_ross", "middle_ross", "upper_ross", and "upper_ross2", while the metadata has site names in the format "Lower Ross", "Middle Ross", "Upper Ross", and "Upper Ross 2". These must be made consistent before I can join the datasets together.

### Clean and standardised site names and convert character to factor variables
```{r fix-site-names}
catch_data <- catch_data |>
  mutate(
    site = str_to_lower(str_trim(site)),
    site = str_replace_all(site, " ", "_")) |> 
  # Convert all chr to fct
  mutate(across(where(is.character), as_factor)
  )

metadata <- metadata |>
  mutate(
    site_name = str_to_lower(str_trim(site_name)),
    site_name = str_replace_all(site_name, " ", "_")) |> 
  # Convert all chr to fct
  mutate(across(where(is.character), as_factor)
  )
```

# Sanity check
```{r recheck-site-names}
catch_data |> 
  str()
 catch_data |> 
   distinct(site)
str(metadata)

```
OK site names fixed in both now (but note the vars themselves still have different names - site vs site_name). We can fix this when we join the datasets together.
Next we need to do something about the species names. Can use the taxonomic dictionary to get scientific names for the species. 

###  Load Taxonomic Dictionary
```{r load-species-dict}
species_dict <- read_csv("../data/workshop2/species_dictionary.csv")
# Check
head(species_dict)
```
Must reference common names in catch log to those in the taxonomic dictionary before we do a join.

```{r check-species-names}
catch_data |> 
  distinct(species)
# Can see these are different to the dictionary
species_dict |>
  distinct(common_name)
```
Only difference is between mangrove_jack and MANGROVE JACK. We can fix this by standardising the species names in the catch log to match those in the dictionary.

###  Standardise species (common name) variable in the catch log to match dictionary
```{r fix-species-names}
catch_data <- catch_data |>
  mutate(
    species = str_to_lower(str_trim(species)),
    species = str_replace_all(species, " ", "_")
  )
```

Still must create scientific_name variables

### Create new variable ("scientific_name") in catch log by joining with taxonomic dictionary 
```{r join-species-dict}
catch_data <- catch_data |>
  mutate(common_name = species) |>
  left_join(species_dict, by = join_by(common_name))
```

# Sanity check: site names + scientific names
```{r}
glimpse(catch_data)
```
Looks cleaner, but must still add metadata in, and must incorporate dataset with sonde/sensor data.

# Load the sonde data
(Code for this is written in separate script which we can source from here)
```{r load-sonde}

source(here::here("code/diw_sensor.R"))
```

#### Notes on sensor data
Sensor data includes 15-minute sonde data for the same 4 sites and 30 days of sampling. The sourced script above turned this into daily data and removed outliers. There are three different environmental variables measured: temperature, salinity, and turbidity. We could investigate the effects of all three on fish species abundance or pick one... Let's remember that the goal of this exercise is to produce a publication-ready visualisation of the estuary's ecology.

The sensor data is now in a dataframe called sensor_daily, which has the following variables: site, date, mean_temp, mean_salinity, and mean_turbidity. We can join this with the catch data and metadata to create a final dataset for analysis.

### Join catch data with metadata and sensor data
```{r join-all}
catch_data_final <- catch_data |>
  left_join(metadata, by = join_by(site == site_name)) |>
  left_join(sensor_daily, by = join_by(site, date))
```

# Sanity check
```{r check-final-join}
glimpse(catch_data_final)
```
At first glance it looks good. However, note that there are only 427 rows of data. This is less than the expected 480 rows (4 species x 4 sites x 30 surveys). This discrepancy is likely due to some days where there were zero counts of certain species at different sites. We need to wrangle these into a complete dataset with 480 rows, where the zero counts are explicitly represented as 0 rather than NA. This is important for our subsequent analyses and visualisations, as we want to be able to distinguish between true absences (0) and missing data (NA)

# Wrangle final dataset

```{r wrangle-final-zeros}
catch_data_final <- 
catch_data_final |> 
  complete(
    nesting(site, date, 
           # lat, lon, 
            zone, mean_temp, mean_salinity, mean_turbidity),
    scientific_name
  ) |>
  mutate(
    # Convert the resulting NA counts to 0
    count = coalesce(count, 0)
  )

catch_data_final
```
Perfect, we now have 480 rows. Now, we can shift gears and start thinking about how we want to analyze and interpret this data. We can start by looking at the environmental variables (salinity, turbidity, temperature) across the different zones (downstream, midstream, upstream), and then we can look at the fish abundance patterns across these zones.

#### Salinity
```{r}
# Table
sal_sum <- 
  catch_data_final |>
  group_by(zone) |>
  summarise(
   sem_salinity = sd(mean_salinity) / sqrt(n()),
    sd_salinity = sd(mean_salinity, na.rm = TRUE),
    median_salinity = median(mean_salinity, na.rm = TRUE),
   mean_salinity = mean(mean_salinity, na.rm = TRUE),
    .groups = "drop"
  )
sal_sum

# Create a figure theme to use for all figures for consistency and professionalism
marine_theme <- theme_classic() +
theme(
text = element_text(size = 14, colour = "black"),
axis.title = element_text(size = 14, face = "bold"),
axis.text = element_text(size = 12),
plot.title = element_text(size = 16, face = "bold"),
plot.subtitle = element_text(size = 13),
legend.title = element_text(size = 13),
legend.text = element_text(size = 12)
)

#| fig-cap: "Figure 1. Median Salinity by Zone with Standard Deviation Error Bars"
#| subtitle: "Data from May 2026, Ross River Estuary System, Townsville, QLD, Australia"
#| fig-height: 6
#| fig-width: 8
# Visualize
sal_sum |> 
  ggplot(aes(x = zone, y = median_salinity)) +
  geom_point() +
  geom_errorbar(aes(ymin = median_salinity - sd_salinity, ymax = median_salinity + sd_salinity), width = 0.2) +
  marine_theme +
  labs(#title = "Median Salinity by Zone with Standard Deviation Error Bars",
       x = "Zone",
       y = "Median Salinity (ppt)")
```
# Interpretation
There is a clear difference in salinity across the zones, with the downstream zone having the highest salinity, and the upstream zone having the lowest. This is consistent with our expectations for an estuarine system, where salinity typically decreases as you move upstream from the ocean.

### Turbidity
```{r}
turb_sum <- 
  catch_data_final |>
  group_by(zone) |>
  summarise(
    sd_turbidity = sd(mean_turbidity, na.rm = TRUE),
    median_turbidity = median(mean_turbidity, na.rm = TRUE),
    mean_turbidity = mean(mean_turbidity, na.rm = TRUE),
    .groups = "drop"
  )

#| fig-cap: "Figure 2. Median Turbidity by Zone with Standard Deviation Error Bars"
#| subtitle: "Data from May 2026, Ross River Estuary System, Townsville, QLD, Australia"
#| fig-height: 6
#| fig-width: 8
turb_sum |>
  ggplot(aes(x = zone, y = median_turbidity)) +
  geom_point() +
  geom_errorbar(aes(ymin = median_turbidity - sd_turbidity, ymax = median_turbidity + sd_turbidity), width = 0.2) +
  marine_theme +
  labs(
       x = "Zone",
       y = "Median Turbidity (NTU)")
```
# Interpretation
Median turbidity is highest in the midstream zone, and lowest in the downstream zone. This could be due to a number of factors, including sediment input from upstream sources, or differences in water flow and mixing across the zones. However overall, the differences in turbidity across the zones are not as pronounced as those for salinity, and the standard deviations are relatively large (range: 2.1-4.8 NTU), indicating high variability in turbidity within each zone.


### Temperature
```{r}
temp_sum <- 
  catch_data_final |>
  group_by(zone) |>
  summarise(
    sd_temperature = sd(mean_temp, na.rm = TRUE),
    median_temperature = median(mean_temp, na.rm = TRUE),
    mean_temperature = mean(mean_temp, na.rm = TRUE),
    .groups = "drop"
  )

#| fig-cap: "Figure 3. Median Temperature by Zone with Standard Deviation Error Bars"
#| subtitle: "Data from May 2026, Ross River Estuary System, Townsville, QLD, Australia"
#| fig-height: 6
#| fig-width: 8
temp_sum |>
  ggplot(aes(x = zone, y = median_temperature)) +
  geom_point() +
  geom_errorbar(aes(ymin = median_temperature - sd_temperature, ymax = median_temperature + sd_temperature), width = 0.2) +
  marine_theme +
  labs(
       x = "Zone",
       y = "Median Temperature (°C)")

```
# Interpretation
Median temperature is highest in the upstream zone, and lowest in the downstream zone (excluding the marine zone). This is consistent with our expectations for an estuarine system, where water temperature typically increases as you move upstream from the ocean. However, the differences in temperature across the zones are not as pronounced as those for salinity, and the standard deviations are relatively large (range: 0.8-1.2 °C), indicating high variability in temperature within each zone.

Because temperature and turbidity are not as strongly differentiated across the zones as salinity, I will focus on salinity as my main environmental variable of interest in my subsequent analyses and visualisations (though I will keep in mind the differences in turbidity and temperature across the zones, which could also be influencing fish abundance patterns).

#### Publish-ready figures
```{r}
#| fig-cap: "Figure 4.  Median Salinity of the Ross River Estuary System, Townsville, QLD, Australia, by Zone with Standard Deviation Error Bars. Data from May 2026"
#| fig-height: 6
#| fig-width: 8

sal_sum |> 
  ggplot(aes(x = zone, y = median_salinity)) +
  geom_point() +
  geom_errorbar(aes(ymin = median_salinity - sd_salinity, ymax = median_salinity + sd_salinity), width = 0.2) +
  marine_theme +
  labs(
       x = "Zone",
       y = "Median Salinity (ppt)")
  
```

# Fish abundance

# Calculate summary stats 
```{r}
fish_stats <- catch_data_final |>
  group_by(zone, scientific_name) |>
  summarise(
    mean_catch = mean(count),
    sem_catch = sd(count) / sqrt(n()),
    .groups = "drop"
  )
fish_stats
```

# Visualization

```{r}
#| fig-cap: "Figure 5. Predatory Fish Abundance across Estuarine Salinity Gradients in the Ross River Estuary System, Townsville, QLD, Australia, May 2026. Mean abundance (n day^-1) with standard error bars for each species across the three zones (downstream, midstream, upstream). Points represent individual survey counts (n = 30 per species per zone)."
#| subtitle: "The y-axis is on a log scale to better visualize the differences in abundance across species and zones."
#| fig-height: 6
#| fig-width: 10

earth_cols <- c(
"#D8B365", # sand
"#8C6D46", # brown
"#5A7D5A", # sage
"#4F7A8A" # muted blue-grey
)

earth_cols_dark <- c(
"#B08E45",
"#6B5231",
"#416041",
"#365765"
)

fish_stats |> 
  ggplot(aes(x = zone, 
             y = mean_catch, 
             fill = scientific_name, 
             colour = scientific_name)) +
  geom_line(group = 1) +
  # add in points from master data
  geom_point(
    data = catch_data_final, 
    aes(x = zone, 
        y = count, 
        colour = scientific_name),
    position = position_jitter(width = 0.2), 
    alpha = 0.4, 
    size = 1.5, 
    inherit.aes = FALSE) + 
  geom_point(
    position = "dodge", 
    size = 5, 
    pch = 23,
    colour = "black") +
  scale_fill_manual(values = earth_cols) +
  scale_colour_manual(values = earth_cols_dark) +
  # SE pretty narrow, potentially not worth plotting but let's viewer see that included
  geom_errorbar(aes(ymin = mean_catch - sem_catch, ymax = mean_catch + sem_catch), 
                position = position_dodge(0.9), width = 0.2) + 
  facet_wrap(~ scientific_name, scales = "fixed") +
  marine_theme +
  theme(
    strip.text = element_text(face = "italic", size = 12, colour = "black"), # spp names in italic
   # legend.position = "bottom"
    legend.position = "none" # legend unnecessary
  ) +
  labs(
    x = "Spatial Zone",
    # scientific label for y axis of n day -1 use regex to get superscript
    y = expression("Abundance (n day"^-1*")"),
  #  y = "Daily abundance (n day^-1)"
    y = "Abundance (per day)"
  )
```





