---
title: "workshop_04_copepodes"
author: "Kendra"
date: "2026-10-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# Load into R Library
library(tidyverse)
library(sf)
library(terra)
library(leaflet)
library(tmap)
library(dichromat)
library(here)

# load copepode data 
dat <- read_csv(here("data", "workshop4", "copepods_raw.csv")) 
dat
```

silk_id shows the unique identifier for each copepod sample, while the other columns provide information about the sampling location, date, and various environmental parameters measured at the time of collection.

For processing, silks are divided into segments based on the `silk_id`. Each segment represents a unique sampling event, and the data can be grouped accordingly for analysis.

Check coordinate reference system (CRS) of the copepod data. The coordinates are in decimal degrees, which is a common format for geographic data.

```{r}
st_crs(dat)
```

## Data exploration

Check coordinates

```{r}
ggplot(dat) + 
  aes(x = longitude, y = latitude, color = richness_raw) +
  geom_point()
```

## Check other patterns in the data
Visualise species richness in a non-spatial way

```{r}
ggplot(dat, aes(x = latitude, y = richness_raw)) +
  stat_smooth() +
  geom_point()
  labs(x = "Latitude", y = "Species Richness") +
  theme_minimal()
```

From this we can see there is an unatural change in the data pattern at around latitude -44. 

Do the same thing with longitude
```{r}
ggplot(dat, aes(x = longitude, y = richness_raw)) +
  stat_smooth() +
  geom_point() +
  labs(x = "Longitude", y = "Species Richness") +
  theme_minimal()
```

```{r}
# Build a simple summary table to summarise the variables in the dataset

overview <- dat %>%
  summarise(
    n_samples = n(),
    n_unique_silk_id = n_distinct(silk_id),
    min_latitude = min(latitude, na.rm = TRUE),
    max_latitude = max(latitude, na.rm = TRUE),
    min_longitude = min(longitude, na.rm = TRUE),
    max_longitude = max(longitude, na.rm = TRUE),
    mean_richness_raw = mean(richness_raw, na.rm = TRUE),
    sd_richness_raw = sd(richness_raw, na.rm = TRUE)
  )
overview
```

## Maps

```{r}
# Convert point data into spatially referenced data using sf package

sdat <- st_as_sf(
dat,
coords = c("longitude", "latitude"),
crs = 4326
)

sdat
```
```{r include = FALSE}

crs5126 <- st_crs(5126)
crs5126 # look at the whole CRS
crs5126$Name # pull out just the name of the crs

crs5126$wkt # pull out the whole WKT of the crs
```

Must decide how to project the data. Projections are defined by a projected coordinate reference system (CRS) that transforms the curved surface of the Earth into a flat plane. The choice of projection can affect the accuracy and interpretation of spatial data, especially for analyses involving distances, areas, and spatial relationships.

```{r}
sdat
```

## Cartography

```{r}
plot(sdat["richness_raw"])
```
Here we have only plotted the richness column. If we used plot(sdat) it would create a panel for every variable in our dataframe. In sf, we can use square brackets ["richness_raw"] to select a single variable.

# Thematic maps for communication

```{r}
# use tmap to create a basic map

tm_shape(sdat) + 
  tm_dots(col = "richness_raw")
```
```{r}
# Improve color scheme and aesthetics of the map

tm1 <- tm_shape(sdat) + 
  tm_dots(col = "richness_raw", 
          palette = "-Blues", 
          size = 0.3, 
          alpha = 0.7) 

tm1

# Save image to outputs
tmap_save(tm1, filename = here("outputs/workshop4", "Richness-map.png"), 
          width = 3000, height = 3000)

```

## Mapping spatial polygons as layers

We will load in a map of Australia and a map of Australia’s continental shelf using tmap to add these layers.

```{r}
aus <- st_read(
  here("data", "workshop4", "spatial-data", "Aussie", "Aussie.shp")
)

shelf <- st_read(
  here("data", "workshop4", "spatial-data", "aus_shelf", "aus_shelf.shp")
)

aus

shelf
```
## Mapping polygons
```{r}
tm_shape(shelf, bbox = sdat) +
  tm_polygons(col = "lightblue") +
  tm_shape(aus) +
  tm_polygons(col = "gray") +
  tm_shape(sdat) +
  tm_dots()
```
```{r}
# Map style 
tmap_style("white")

tm2 <- tm_shape(shelf, bbox = sdat) +
  tm_polygons(col = "lightblue") +
  tm_shape(aus) +
  tm_polygons(col = "grey85") +
  tm_shape(sdat) +
  tm_dots(
    col = "richness_raw",
    palette = "-YlGnBu",
    size = 0.1
  ) +
  tm_layout(
    legend.position = c("left", "top"),
    frame = TRUE
  )

tm2

tm2 <- tmap_save(tm2, filename = here("outputs/workshop4", "Richness-map-aus-shelf.png"), 
          width = 3000, height = 3000)

```

## tmp vignette
```{r}
vignette('tmap-getstarted')

tmap_mode("view")

data("World")
tm_shape(World) +
  tm_polygons("HPI")
```
```{r}
data(World, metro, rivers, land)

tmap_mode("plot")

tm_shape(land) +
  tm_raster("elevation", palette = terrain.colors(10)) +
tm_shape(World) +
    tm_borders("white", lwd = .5) +
    tm_text("iso_a3", size = "AREA") +
tm_shape(metro) +
    tm_symbols(col = "red", size = "pop2020", scale = .5) +
tm_legend(show = FALSE)
```
## Facets
```{r}
tmap_mode("view")
tm_shape(World) +
    tm_polygons(c("HPI", "economy")) +
    tm_facets(sync = TRUE, ncol = 2)
```
```{r}
tmap_mode("plot")
## tmap mode set to plotting

data(NLD_muni)

NLD_muni$perc_men <- NLD_muni$pop_men / NLD_muni$population * 100

tm_shape(NLD_muni) +
    tm_polygons("perc_men", palette = "RdYlBu") +
    tm_facets(by = "province")
```
```{r}
tmap_mode("plot")
## tmap mode set to plotting

data(NLD_muni)
tm1 <- tm_shape(NLD_muni) + tm_polygons("population", convert2density = TRUE)
tm2 <- tm_shape(NLD_muni) + tm_bubbles(size = "population")

tmap_arrange(tm1, tm2)
```
## Style
```{r}
tmap_mode("plot")
## tmap mode set to plotting

tm_shape(World) +
    tm_polygons("HPI") +
tm_layout(bg.color = "skyblue", inner.margins = c(0, .02, .02, .02))
```
## Setting options
```{r}
tmap_options(bg.color = "black", legend.text.color = "white")

tm_shape(World) +
    tm_polygons("HPI", legend.title = "Happy Planet Index")

```
```{r}
tmap_style("classic")
## tmap style set to "classic"
## other available styles are: "white", "gray", "natural", "cobalt", "col_blind", "albatross", "beaver", "bw", "watercolor"

tm_shape(World) +
    tm_polygons("HPI", legend.title = "Happy Planet Index")

tmap_options_reset()
```
## Exporting
```{r}
tm3 <- tm_shape(World) +
    tm_polygons("HPI", legend.title = "Happy Planet Index")

## save an image ("plot" mode)
tmap_save(tm3, filename = here("outputs/workshop4","vignette_world_map.png"),          width = 3000, height = 3000)
```
```{r}
tmap_save(tm3, filename = here("outputs/workshop4","vignette_world_map.html"))
```
## Quick Thematic Map
```{r}
qtm(World, fill = "HPI", fill.palette = "RdYlGn")
```
```{r}
tmap_tip()
```


