---
title: "workshop_04_copepodes"
author: "Kendra"
date: "2026-10-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# Load into R Library
library(tidyverse)
library(sf)
library(terra)
library(leaflet)
library(tmap)
library(dichromat)
library(here)

# load copepode data 
dat <- read_csv(here("data", "workshop4","copepods", "copepods_raw.csv")) 
dat
```

silk_id shows the unique identifier for each copepod sample, while the other columns provide information about the sampling location, date, and various environmental parameters measured at the time of collection.

For processing, silks are divided into segments based on the `silk_id`. Each segment represents a unique sampling event, and the data can be grouped accordingly for analysis.

Check coordinate reference system (CRS) of the copepod data. The coordinates are in decimal degrees, which is a common format for geographic data.

```{r}
st_crs(dat)
```

## Data exploration

Check coordinates

```{r}
ggplot(dat) + 
  aes(x = longitude, y = latitude, color = richness_raw) +
  geom_point()
```

## Check other patterns in the data
Visualise species richness in a non-spatial way

```{r}
ggplot(dat, aes(x = latitude, y = richness_raw)) +
  stat_smooth() +
  geom_point()
  labs(x = "Latitude", y = "Species Richness") +
  theme_minimal()
```

From this we can see there is an unatural change in the data pattern at around latitude -44. 

Do the same thing with longitude
```{r}
ggplot(dat, aes(x = longitude, y = richness_raw)) +
  stat_smooth() +
  geom_point() +
  labs(x = "Longitude", y = "Species Richness") +
  theme_minimal()
```

```{r}
# Build a simple summary table to summarise the variables in the dataset

overview <- dat %>%
  summarise(
    n_samples = n(),
    n_unique_silk_id = n_distinct(silk_id),
    min_latitude = min(latitude, na.rm = TRUE),
    max_latitude = max(latitude, na.rm = TRUE),
    min_longitude = min(longitude, na.rm = TRUE),
    max_longitude = max(longitude, na.rm = TRUE),
    mean_richness_raw = mean(richness_raw, na.rm = TRUE),
    sd_richness_raw = sd(richness_raw, na.rm = TRUE)
  )
overview
```

## Maps

```{r}
# Convert point data into spatially referenced data using sf package

sdat <- st_as_sf(
dat,
coords = c("longitude", "latitude"),
crs = 4326
)

sdat
```
```{r include = FALSE}

crs5126 <- st_crs(5126)
crs5126 # look at the whole CRS
crs5126$Name # pull out just the name of the crs

crs5126$wkt # pull out the whole WKT of the crs
```

Must decide how to project the data. Projections are defined by a projected coordinate reference system (CRS) that transforms the curved surface of the Earth into a flat plane. The choice of projection can affect the accuracy and interpretation of spatial data, especially for analyses involving distances, areas, and spatial relationships.

```{r}
sdat
```

## Cartography

```{r}
plot(sdat["richness_raw"])
```
Here we have only plotted the richness column. If we used plot(sdat) it would create a panel for every variable in our dataframe. In sf, we can use square brackets ["richness_raw"] to select a single variable.

# Thematic maps for communication

```{r}
# use tmap to create a basic map

tm_shape(sdat) + 
  tm_dots(col = "richness_raw")
```
```{r}
# Improve color scheme and aesthetics of the map

tm1 <- tm_shape(sdat) + 
  tm_dots(col = "richness_raw", 
          palette = "-Blues", 
          size = 0.3, 
          alpha = 0.7) 

tm1

# Save image to outputs
tmap_save(tm1, filename = here("outputs/workshop4", "Richness-map.png"), 
          width = 3000, height = 3000)

```

## Mapping spatial polygons as layers

We will load in a map of Australia and a map of Australia’s continental shelf using tmap to add these layers.

```{r}
aus <- st_read(
  here("data", "workshop4", "copepods","spatial-data", "Aussie", "Aussie.shp")
)

shelf <- st_read(
  here("data", "workshop4", "copepods", "spatial-data", "aus_shelf", "aus_shelf.shp")
)

aus

shelf
```
## Mapping polygons
```{r}
tm_shape(shelf, bbox = sdat) +
  tm_polygons(col = "lightblue") +
  tm_shape(aus) +
  tm_polygons(col = "gray") +
  tm_shape(sdat) +
  tm_dots()
```
```{r}
# Map style 
tmap_style("white")

tm2 <- tm_shape(shelf, bbox = sdat) +
  tm_polygons(col = "lightblue") +
  tm_shape(aus) +
  tm_polygons(col = "grey85") +
  tm_shape(sdat) +
  tm_dots(
    col = "richness_raw",
    palette = "-YlGnBu",
    size = 0.1
  ) +
  tm_layout(
    legend.position = c("left", "top"),
    frame = TRUE
  )

tm2

tm2 <- tmap_save(tm2, filename = here("outputs/workshop4", "Richness-map-aus-shelf.png"), 
          width = 3000, height = 3000)

```

## tmp vignette
```{r}
vignette('tmap-getstarted')

tmap_mode("view")

data("World")
tm_shape(World) +
  tm_polygons("HPI")
```
```{r}
data(World, metro, rivers, land)

tmap_mode("plot")

tm_shape(land) +
  tm_raster("elevation", palette = terrain.colors(10)) +
tm_shape(World) +
    tm_borders("white", lwd = .5) +
    tm_text("iso_a3", size = "AREA") +
tm_shape(metro) +
    tm_symbols(col = "red", size = "pop2020", scale = .5) +
tm_legend(show = FALSE)
```
## Facets
```{r}
tmap_mode("view")
tm_shape(World) +
    tm_polygons(c("HPI", "economy")) +
    tm_facets(sync = TRUE, ncol = 2)
```
```{r}
tmap_mode("plot")
## tmap mode set to plotting

data(NLD_muni)

NLD_muni$perc_men <- NLD_muni$pop_men / NLD_muni$population * 100

tm_shape(NLD_muni) +
    tm_polygons("perc_men", palette = "RdYlBu") +
    tm_facets(by = "province")
```
```{r}
tmap_mode("plot")
## tmap mode set to plotting

data(NLD_muni)
tm1 <- tm_shape(NLD_muni) + tm_polygons("population", convert2density = TRUE)
tm2 <- tm_shape(NLD_muni) + tm_bubbles(size = "population")

tmap_arrange(tm1, tm2)
```
## Style
```{r}
tmap_mode("plot")
## tmap mode set to plotting

tm_shape(World) +
    tm_polygons("HPI") +
tm_layout(bg.color = "skyblue", inner.margins = c(0, .02, .02, .02))
```
## Setting options
```{r}
tmap_options(bg.color = "black", legend.text.color = "white")

tm_shape(World) +
    tm_polygons("HPI", legend.title = "Happy Planet Index")

```
```{r}
tmap_style("classic")
## tmap style set to "classic"
## other available styles are: "white", "gray", "natural", "cobalt", "col_blind", "albatross", "beaver", "bw", "watercolor"

tm_shape(World) +
    tm_polygons("HPI", legend.title = "Happy Planet Index")

tmap_options_reset()
```
## Exporting
```{r}
tm3 <- tm_shape(World) +
    tm_polygons("HPI", legend.title = "Happy Planet Index")

## save an image ("plot" mode)
tmap_save(tm3, filename = here("outputs/workshop4","vignette_world_map.png"),          width = 3000, height = 3000)
```
```{r}
tmap_save(tm3, filename = here("outputs/workshop4","vignette_world_map.html"))
```
## Quick Thematic Map
```{r}
qtm(World, fill = "HPI", fill.palette = "RdYlGn")
```
```{r}
tmap_tip()
```


## Tutorial with data for tmap ##

Load data and packages for this tutorial
```{r}
# for loading our data
library(raster)
library(readr)
library(readxl)
library(sf)
# for datasets
library(maps)
library(spData)
# for creating animations
library(magick)
# for plotting
library(grid)
library(tmap)
library(viridis)
```




```{r}
# load shapefile for bavaria
bavaria <- read_sf(
  here("data", "workshop4", "chapter2", "bavaria.shp")
)

# nicer colnames
colnames(bavaria) <- c(
  "place", "type", "gdp_per_capita", "mean_age",
  "pop_density", "unemployment_rate",
  "employment_rate", "household_income",
  "students", "pop_development", "geometry")

  
# load raster file for europe
europe_raster <- raster(
  here("data", "workshop4", "chapter2", "elevation1x1_new.tif")
)

# load shapefile for world
world_shape <- read_sf(
  here("data", "workshop4", "chapter2", "ne_50m_admin_0_countries.shp")
)

# keep only europe
europe_shape <- world_shape[
  world_shape$CONTINENT == "Europe", ]

# keep only cities with at least 1 million inhabitants
cities <- world.cities[world.cities$pop >= 1000000, ]

# turn it into an sf object
cities <- cities %>%
  st_as_sf(coords = c("long", "lat"), crs = 4326) %>%
  st_cast("POINT")

# keep only cities that are in europe
cities <- st_intersection(cities, st_union(europe_shape))

# turn europe into a MULTILINESTRING
europe_shape <- st_cast(europe_shape, "MULTILINESTRING")

# load german municipalities
communities <- read_sf(
  here("data", "workshop4", "chapter2", "gmd_ex.shp")
)

# keep only Rosenheim
rosenheim <- communities[
  communities$BEZ_KRS == "Rosenheim", ]

# load honey production data
honey_csv <- read_csv(
  here("data", "workshop4", "chapter2", "honeyproduction.csv")
)
colnames(honey_sf)[3:8] <- c(
  "Number_of_colonies", "Yield_per_colony",
  "Total_production", "Stocks","Price_per_lb",
  "Value_of_production"
  )

# load state abbreviations
abbrev <- read_xlsx(
  here("data", "workshop4", "chapter2", "abbrev.xlsx")
)

# load honey shapefile
honey_sf <- read_sf(
  here("data", "workshop4", "chapter2", "honey.shp")
)
```
## Static maps with tmap
```{r}
tm_shape(bavaria) +
  tm_fill()
tm_shape(bavaria) +
  tm_borders()
tm_shape(bavaria) +
  tm_polygons()

# save map as an object

map_europe <- tm_shape(europe_raster) +
  tm_raster()
class(map_europe)
```


# Use of several layers
```{r}
map_europe2 <- map_europe +
  tm_shape(europe_shape) +
  tm_lines(alpha = 0.3)
map_europe3 <- map_europe2  +
  tm_shape(cities) +
  tm_dots(size = 0.3)

map_europe2
map_europe3
```


## Use of various arguments
```{r message=FALSE, warning=FALSE}
bav1 <- tm_shape(bavaria) +
  tm_fill(col = "aquamarine")
bav2 <- tm_shape(bavaria) +
  tm_fill(col = "aquamarine", alpha = 0.5)
bav3 <- tm_shape(bavaria) +
  tm_polygons(col = "aquamarine", border.col = "darkolivegreen")
bav4 <- tm_shape(bavaria) +
  tm_borders(lwd = 2)
bav5 <- tm_shape(bavaria) +
  tm_borders(lty = 4)
bav6 <- tm_shape(bavaria) +
  tm_polygons(col = "#E2E2E2", border.alpha = 0.5, lwd = 3)
all_bavs <- tmap_arrange(bav1, bav1, bav3, bav4, bav5, bav6, ncol = 3)

all_bavs
```

```{r}
tm_shape(bavaria) +
  tm_polygons(col = "pop_development", midpoint = 0)
```

```{r}
tm_shape(bavaria) +
  tm_polygons(col = "mean_age")
tm_shape(bavaria) +
  tm_polygons(col = "mean_age", breaks = c(40, 42, 46, 54))
tm_shape(bavaria) +
  tm_polygons(col = "mean_age", n = 3)
```

```{r}
tm_shape(bavaria) +
  tm_polygons(col = "unemployment_rate", palette = "viridis")
tm_shape(bavaria) +
  tm_polygons(col = "unemployment_rate", palette = "-viridis")
```



Map of Bavaria with additional elements

```{r}
tm_shape(bavaria) +
  tm_polygons() +
  tm_layout(title = "Bavaria", bg.color = "#228B22")
tm_shape(bavaria) +
  tm_polygons() +
  tm_layout(scale = 3, frame = FALSE)
tm_shape(bavaria) +
  tm_polygons(col = "type") +
  tm_layout(frame.lwd = 3, legend.position = c("left", "bottom"))
tm_shape(bavaria) +
  tm_polygons(col = "type") +
  tm_layout(inner.margins = 0.2, legend.show = FALSE)
```

Some layout settings

```{r}
tm_shape(bavaria) +
  tm_polygons() +
  tm_layout(title = "Bavaria", bg.color = "#228B22")
tm_shape(bavaria) +
  tm_polygons() +
  tm_layout(scale = 3, frame = FALSE)
tm_shape(bavaria) +
  tm_polygons(col = "type") +
  tm_layout(frame.lwd = 3, legend.position = c("left", "bottom"))
tm_shape(bavaria) +
  tm_polygons(col = "type") +
  tm_layout(inner.margins = 0.2, legend.show = FALSE)
```

Classic and color-blind style
```{r}
tm_shape(bavaria) +
  tm_polygons(col = "type") +
  tm_style("classic")
tm_shape(bavaria) +
  tm_polygons(col = "type") +
  tm_style("col_blind")
```

```{r}
us_states2163 <- st_transform(us_states, 2163)

us_states_range <- st_bbox(us_states2163)[4] - st_bbox(us_states2163)[2]
hawaii_range <- st_bbox(hawaii)[4] - st_bbox(hawaii)[2]
alaska_range <- st_bbox(alaska)[4] - st_bbox(alaska)[2]
us_states_hawaii_ratio <- hawaii_range / us_states_range
us_states_alaska_ratio <- alaska_range / us_states_range
```

```{r}
us_states_map <- tm_shape(us_states2163) +
  tm_polygons(
    col = "total_pop_15",
    breaks = c(
      0, 5000000, 10000000, 15000000, 20000000,
      25000000, 30000000, 35000000, 40000000
      ),
    title = "Population 2015 in millions",
    pal = c("#E1F5C4", "#EDE574", "#F9D423", "#FC913A", "#FF4E50"),
    labels = c(
      "0-5", "5-10", "10-15", "15-20", "20-25",
      "25-30", "30-35", "35-40"
      ),
    legend.is.portrait = FALSE) +
  tm_layout(
    frame = FALSE,
    legend.outside = TRUE,
    legend.outside.position = "bottom"
    )
hawaii_map <- tm_shape(hawaii) +
  tm_polygons(
    col = "total_pop_15",
    breaks = c(
      0, 5000000, 10000000, 15000000, 20000000,
      25000000, 30000000, 35000000, 40000000
      ),
    pal = c("#E1F5C4", "#EDE574", "#F9D423", "#FC913A", "#FF4E50")) +
  tm_layout(
    title = "Hawaii",
    frame = FALSE,
    bg.color = NA,
    legend.show = FALSE,
    title.position = c("LEFT", "BOTTOM")
    )
alaska_map <- tm_shape(alaska) +
  tm_polygons(
    col = "total_pop_15",
    breaks = c(
      0, 5000000, 10000000, 15000000, 20000000,
      25000000, 30000000, 35000000, 40000000),
    pal = c("#E1F5C4", "#EDE574", "#F9D423", "#FC913A", "#FF4E50")) +
  tm_layout(
    title = "Alaska",
    frame = FALSE,
    bg.color = NA,
    legend.show = FALSE,
    title.position = c("LEFT", "TOP")
    )
```
```{r}
pushViewport(
  viewport(
    layout = grid.layout(
      2, 1, heights = unit(c(us_states_alaska_ratio, 1), "null")
      )
    )
  )
print(
  alaska_map,
  vp = viewport(layout.pos.row = 1)
  )
print(
  us_states_map,
  vp = viewport(layout.pos.row = 2)
  )
print(
  hawaii_map,
  vp = viewport(
    x = 0.1, y = 0.07,
    height = us_states_hawaii_ratio / sum(c(us_states_alaska_ratio, 1))
    )
  )
grid.lines(x = c(0, 1), y = c(0.58, 0.58), gp = gpar(lty = 2))
grid.lines(x = c(0, 0.2), y = c(0.33, 0), gp = gpar(lty = 2))
```


```{r}
# select the years starting in 2009
honey0912 <- honey_csv[honey_csv$year > 2008, ]
# bind missing data to the original data
honey0912 <- rbind(
  honey0912,
  # create an empty matrix for all the states that are missing
  matrix(
    NA,
    ncol = ncol(honey0912),
    nrow = 4 * length(
      unique(abbrev$state)[!unique(abbrev$state) %in% unique(honey0912$state)]
      ),
    dimnames = list(
      c(
        seq_len(
          4 * length(
            unique(abbrev$state)[!unique(abbrev$state) %in% unique(honey0912$state)])
          )
        ),
      colnames(honey0912)
      )
    )
  )
# add the missing state names
honey0912[
  is.na(honey0912$state), ]$state <- sort(
    rep(unique(abbrev$state)[!unique(abbrev$state) %in% unique(honey0912$state)], 4)
    )
# add the year for the missing states
honey0912[is.na(honey0912$year), ]$year <- rep(2009:2012, 10)
# merge the honey dataset with the state abbreviations
honey0912 <-  merge.data.frame(honey0912, abbrev, by = "state")
# change the colnames of our two datasets for merging
colnames(honey0912)[c(1, 9)] <- c("short", "State")
# use the us_states data from spData
colnames(us_states)[2] <- "State"
# remove the district of columbia from the us_states dataset from spData
us_states <- us_states[us_states$State != "District of Columbia", ]
# merge the data together
honey0912 <- merge(honey0912, us_states[, c(2, 7)], by = "State", all.y = TRUE)
# turn it into an sf object
honey0912 <- st_as_sf(honey0912)
```

```{r}
tm_shape(honey0912) +
  tm_polygons(
    col = "priceperlb",
    style = "cont",
    pal = viridis(10, direction = -1),
    title = "Price in US$ per lb"
    ) +
  tm_facets(by = "year", ncol = 2) +
  tm_layout(legend.outside.size = 0.2)
```
```{r}
library(gifski)
honey_animation <- tm_shape(honey_sf) +
  tm_polygons(
    col = "Price_per_lb",
    style = "cont",
    pal = viridis(10, direction = -1)
    ) +
  tm_facets(along = "year") +
  tm_layout(legend.position = c("left", "bottom"))
tmap_animation(
  honey_animation, filename = "honey.gif",
  delay = 50, width = 2400, height = 1200
  )
```


Interactive maps with tmap
```{r}
tmap_mode("view")
tm_shape(bavaria) + tm_polygons(col = "pop_development", midpoint = 0)
```

```{r}
tm_shape(bavaria) +
  tm_polygons(col = "pop_development", midpoint = 0) +
  tm_basemap("Stamen.Watercolor")
```


```{r}
accidents <- read_csv(
  here(
    "data",
    "workshop4",
    "chapter2",
    "accidents.csv"
  )
)

accidents <- accidents %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326) %>%
  st_cast("POINT")
```
```{r}
london <- read_sf(
  here(
    "data",
    "workshop4",
    "chapter2",
    "London_Borough_Excluding_MHW.shp"
  )
) %>%
  st_transform(4326)

london_union <- st_union(london)

accidents_m <- st_intersects(
  london_union,
  accidents
)

accidents <- accidents[
  unlist(accidents_m),
]

london <- london[london$NAME %in% c("City of London", "Westminster", "Camden"), ]
london <- st_union(london)
accidents_m <- st_intersects(london, accidents)
accidents <- accidents[unlist(accidents_m), ]

# do some ordering for the legend
accidents$Light_Conditions <- ordered(
  accidents$Light_Conditions,
  levels = c(
    "Daylight",
    "Darkness - lights lit",
    "Darkness - lighting unknown",
    "Darkness - lights unlit",
    "Darkness - no lighting",
    "Data missing or out of range"
    )
  )
# define the map
map_london <- tm_shape(accidents) +
  tm_dots(
    group = "2017",
    col = "Light_Conditions",
    palette = "Dark2",
    popup.vars = TRUE
  ) +
  tm_view(alpha = 1)

map_london
```

## REVISITING MY WORKSHOP 4 MAP ##
```{r}

library(lubridate)

sdat <- sdat %>%
  mutate(
    sample_datetime = dmy_hm(sample_time_utc),
    sample_date = as.Date(sample_datetime),
    sample_year = factor(year(sample_datetime)),
    sample_month = floor_date(sample_datetime, unit = "month")
  )

glimpse(sdat)

range(
  sdat$sample_datetime,
  na.rm = TRUE
)
```
```{r}
tmap_mode("view")

richness_map_interactive <- tm_shape(shelf) +
  tm_polygons(
    col = "#BFE3E8",
    alpha = 0.35,
    border.col = NA,
    group = "Continental shelf"
  ) +

  tm_shape(aus) +
  tm_polygons(
    col = "#E6E2D8",
    border.col = "#666666",
    alpha = 0.8,
    group = "Australia"
  ) +

  tm_shape(sdat) +
  tm_dots(
    col = "richness_raw",
    palette = "-YlGnBu",
    style = "cont",
    size = 0.03,
    alpha = 0.75,
    title = "Copepod richness",
    group = "Samples",
    popup.vars = c(
      "Silk ID" = "silk_id",
      "Segment" = "segment_no",
      "Richness" = "richness_raw",
      "Region" = "region",
      "Route" = "route",
      "Vessel" = "vessel",
      "Sample time" = "sample_time_utc"
    )
  ) +

  tm_view(
    alpha = 1
  )

richness_map_interactive

tmap_save(
richness_map_interactive,
filename = here(
"outputs",
"workshop4",
"copepod-richness-interactive.html"
)
)
```





