---
title: "R for Marine Science: Workshop 1"
author: "Kendra Sturdee"
date: today
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
execute:
  warning: false
  message: false
editor: visual
echo: false
---

# 

```{r}
library(tidyverse)
library(palmerpenguins)
library(here)

#| echo: false
#| source(here::here("code", "workshop_01_r4marinescience.R"))
```

## Visualizing Body Size and Shape

```{r}
#| echo: false
#| 
penguins %>%
  ggplot(aes(x = species,
             y = body_mass_g,
             fill = island)) +
  geom_boxplot(color = "black") +
  labs(
    title = "Penguin Body Mass by Species and Island",
    x = "Species",
    y = "Body Mass (g)",
    fill = "Island"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c(
    "darkgreen",
    "saddlebrown",
    "goldenrod"
  )) +
  theme(
    text = element_text(color = "black"),
    axis.text = element_text(color = "black"),
    axis.title = element_text(color = "black"),
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    panel.grid.minor = element_blank(),
    legend.background = element_rect(
      fill = "grey95",
      color = "grey70"
    ),
    legend.key = element_rect(
      fill = "grey95",
      color = NA
    ),
    legend.position = "right",
    plot.title = element_text(
      face = "bold",
      hjust = 0.5
    )
  )

```

## **Visualizing Mean Trends with Uncertainty**

```{r}
#| echo: false
#| # Pipe directly from aggregation to plotting with error bars

mass_compare_plot <- penguins |>
  group_by(species, island) |>
  summarise(
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    sd_mass = sd(body_mass_g, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |>
  ggplot(aes(x = species, y = mean_mass, colour = island)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_mass - sd_mass, 
                    ymax = mean_mass + sd_mass), 
                width = 0.2) +
  labs(
    title = "Mean Body Mass by Species and Island",
    subtitle = "Error bars represent standard deviation",
    y = "Mean Body Mass (g)",
    x = "Species",
    colour = "Island"
  ) +
  scale_colour_manual(
    values = c(
      "darkgreen",
      "saddlebrown",
      "goldenrod"
    )
  ) + 
  theme_minimal() +
  theme(
    text = element_text(color = "black"),
    axis.text = element_text(color = "black"),
    axis.title = element_text(color = "black"),
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    legend.background = element_rect(
      fill = "grey95",
      color = "grey70"
    ),
    legend.key = element_rect(
      fill = "grey95",
      color = NA
    ),
    legend.position = "right",
    plot.title = element_text(
      face = "bold",
      hjust = 0.5
    ),
    plot.subtitle = element_text(
      hjust = 0.5,
      color = "black"
    )
  )


mass_compare_plot

```

```{r}
mass_compare_plot_se <- penguins |>
  group_by(species, island) |>
  summarise(
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    sd_mass = sd(body_mass_g, na.rm = TRUE),
    n = n(),
    se_mass = sd_mass / sqrt(n),
    .groups = "drop"
  ) |>
  ggplot(aes(x = species, y = mean_mass, colour = island)) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(
      ymin = mean_mass - se_mass,
      ymax = mean_mass + se_mass
    ),
    width = 0.2
  ) +
  labs(
    title = "Mean Body Mass by Species and Island",
    subtitle = "Error bars represent standard error",
    y = "Mean Body Mass (g)",
    x = "Species",
    colour = "Island"
  ) +
  scale_colour_manual(
    values = c(
      "darkgreen",
      "saddlebrown",
      "goldenrod"
    )
  ) +
  theme_minimal() +
  theme(
    text = element_text(color = "black"),
    axis.text = element_text(color = "black"),
    axis.title = element_text(color = "black"),
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    legend.background = element_rect(
      fill = "grey95",
      color = "grey70"
    ),
    legend.key = element_rect(
      fill = "grey95",
      color = NA
    ),
    legend.position = "right",
    plot.title = element_text(
      face = "bold",
      hjust = 0.5
    ),
    plot.subtitle = element_text(
      hjust = 0.5,
      color = "black"
    )
  )

mass_compare_plot_se
```

Using standard error instead of standard deviation returned a similar trend with narrower error bars.

```{r}
penguinoverview <- penguins %>%
  group_by(species, island) %>%
  summarise(
    mean_mass_g = mean(body_mass_g, na.rm = TRUE),
    sd_mass_g = sd(body_mass_g, na.rm = TRUE),
    n = n(),
    se_mass_g = sd_mass_g / sqrt(n),
    .groups = "drop"
  )

view(penguinoverview)
```

Overall, it seems like gentoo penguins are the heaviest by a wide margin, while chinstrap and adelie are more closely related in size.

Evidently, mean body mass differed between penguin species, with Gentoo penguins generally exhibiting larger body masses than Adelie and Chinstrap penguins. There was also some variation between islands, suggesting that location may influence morphological characteristics.

While the standard error and standard deviations plots were generally pretty similar, the former showed more precise estimates of mean body mass, while the latter showed the greater variability among individual penguins within each species-island grouping. It seems that both species and environmental context may contribute to variation in biological traits.
