---
title: "R for Marine Science: Workshop 1"
author: "Kendra Sturdee"
date: today
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
execute:
  warning: false
  message: false
editor: visual
echo: false
---

# Overview

In this workshop, I used the Palmer Penguins dataset to explore patterns in penguin morphology and investigate how body size varies among species and locations using data wrangling and visualisation techniques in R. The penguins data includes morphological measurements of three penguin species (Adelie, Chinstrap, and Gentoo) from islands near Palmer Station, Antarctica. Besides body mass, the dataset also includes bill length, bill depth, flipper length, sex, species, and island location.

```{r}
library(tidyverse)
library(palmerpenguins)
library(here)
```

## Visualizing Body Size and Shape

The following figure compares the distribution of body mass between penguin species and islands.

```{r}
#| echo: false

penguins %>%
  ggplot(aes(x = species,
             y = body_mass_g,
             fill = island)) +
  geom_boxplot(color = "black") +
  labs(
    title = "Penguin Body Mass by Species and Island",
    x = "Species",
    y = "Body Mass (g)",
    fill = "Island"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c(
    "darkgreen",
    "saddlebrown",
    "goldenrod"
  )) +
  theme(
    text = element_text(color = "black"),
    axis.text = element_text(color = "black"),
    axis.title = element_text(color = "black"),
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    panel.grid.minor = element_blank(),
    legend.background = element_rect(
      fill = "grey95",
      color = "grey70"
    ),
    legend.key = element_rect(
      fill = "grey95",
      color = NA
    ),
    legend.position = "right",
    plot.title = element_text(
      face = "bold",
      hjust = 0.5
    )
  )

```

## Interpretation

Gentoo penguins had the greatest body mass, while Adelie and Chinstrap penguins tended to be generally lighter. The slight variation between islands may suggest that environmental factors are contributing to differences in body mass.

## Summary Statistics

The mean body mass was then calculated for each penguin species (grouped by island).

```{r}
#| echo: false

penguinoverview <- penguins |>
  group_by(species, island) |>
  summarise(
    mean_mass_g = mean(body_mass_g, na.rm = TRUE),
    sd_mass_g = sd(body_mass_g, na.rm = TRUE),
    n = n(),
    se_mass_g = sd_mass_g / sqrt(n),
    .groups = "drop"
  )

knitr::kable(
  penguinoverview,
  digits = 2,
  caption = "Summary statistics by species and island."
)
```

## **Visualizing Mean Trends with Uncertainty**

The figures below show mean body mass with standard deviation and standard error bars.

```{r}
#| echo: false
#| # Pipe directly from aggregation to plotting with error bars

mass_compare_plot <- penguins |>
  group_by(species, island) |>
  summarise(
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    sd_mass = sd(body_mass_g, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |>
  ggplot(aes(x = species, y = mean_mass, colour = island)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_mass - sd_mass, 
                    ymax = mean_mass + sd_mass), 
                width = 0.2) +
  labs(
    title = "Mean Body Mass by Species and Island",
    subtitle = "Error bars represent standard deviation",
    y = "Mean Body Mass (g)",
    x = "Species",
    colour = "Island"
  ) +
  scale_colour_manual(
    values = c(
      "darkgreen",
      "saddlebrown",
      "goldenrod"
    )
  ) + 
  theme_minimal() +
  theme(
    text = element_text(color = "black"),
    axis.text = element_text(color = "black"),
    axis.title = element_text(color = "black"),
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    legend.background = element_rect(
      fill = "grey95",
      color = "grey70"
    ),
    legend.key = element_rect(
      fill = "grey95",
      color = NA
    ),
    legend.position = "right",
    plot.title = element_text(
      face = "bold",
      hjust = 0.5
    ),
    plot.subtitle = element_text(
      hjust = 0.5,
      color = "black"
    )
  )


mass_compare_plot

```

```{r}
mass_compare_plot_se <- penguins |>
  group_by(species, island) |>
  summarise(
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    sd_mass = sd(body_mass_g, na.rm = TRUE),
    n = n(),
    se_mass = sd_mass / sqrt(n),
    .groups = "drop"
  ) |>
  ggplot(aes(x = species, y = mean_mass, colour = island)) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(
      ymin = mean_mass - se_mass,
      ymax = mean_mass + se_mass
    ),
    width = 0.2
  ) +
  labs(
    title = "Mean Body Mass by Species and Island",
    subtitle = "Error bars represent standard error",
    y = "Mean Body Mass (g)",
    x = "Species",
    colour = "Island"
  ) +
  scale_colour_manual(
    values = c(
      "darkgreen",
      "saddlebrown",
      "goldenrod"
    )
  ) +
  theme_minimal() +
  theme(
    text = element_text(color = "black"),
    axis.text = element_text(color = "black"),
    axis.title = element_text(color = "black"),
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    legend.background = element_rect(
      fill = "grey95",
      color = "grey70"
    ),
    legend.key = element_rect(
      fill = "grey95",
      color = NA
    ),
    legend.position = "right",
    plot.title = element_text(
      face = "bold",
      hjust = 0.5
    ),
    plot.subtitle = element_text(
      hjust = 0.5,
      color = "black"
    )
  )

mass_compare_plot_se
```

## Discussion

While the standard error and standard deviations plots were generally pretty similar, the former showed more precise estimates of mean body mass, while the latter showed the greater variability among individual penguins within each species-island grouping. It seems that both species and environmental context may contribute to variation in biological traits.
